---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, echo = FALSE, message = FALSE, results='hide', warning=FALSE}
# Load libraries
library(plyr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
```

## Loading and preprocessing the data

```{r}
# Unzip the file
unzip(zipfile = 'activity.zip', overwrite = TRUE)
# Read the file
data <- read.csv('./activity.csv') %>%
# Convert date and interval to POSIX dates
  mutate(date = ymd(date),
         interval = as.POSIXct(today() + hm(paste(interval %/% 100, ':',
                                                  interval %% 100,
                                                  sep = '')))) %>%
# Make sure the data is ordered correctly
  arrange(date, interval)
```

## What is mean total number of steps taken per day?

```{r}
# Calculate the sum of steps per day
stepsPerDay <- data %>%
  group_by(date) %>%
  summarize(steps = sum(steps, na.rm = TRUE))
# Calculate the summary data (mean, median)
summaryBeforeImpute <- stepsPerDay %>%
  summarise(mean = round(mean(steps, na.rm = TRUE), 2),
            median = round(median(steps, na.rm = TRUE), 2))
# Plot the histogram
qplot(x = steps, data = stepsPerDay, geom = 'histogram')
```

The mean of the total number of steps taken per day is:
`r format(summaryBeforeImpute$mean)`.  
The median of the total number of steps taken per day is:
`r format(summaryBeforeImpute$median)`.  

## What is the average daily activity pattern?

```{r}
# Calculate the mean of steps per interval
stepsPerInterval <- data %>%
  group_by(interval) %>%
  summarize(steps = mean(steps, na.rm = TRUE))
# Plot the time series
qplot(x = interval, y = steps, data = stepsPerInterval, geom = 'line') +
  scale_x_datetime(breaks = '4 hours', labels = date_format('%H:%M'))
# Keep only the line corresponding to the maximum mean
stepsPerInterval <- stepsPerInterval %>%
  filter(steps == max(steps)) %>%
  mutate(steps = round(steps, 2))
```

The interval with the maximum number of steps is at
`r format(stepsPerInterval$interval, '%H:%M')` (`r format(stepsPerInterval$steps)` steps on
average).

## Imputing missing values

```{r}
# Count the nomber of incomplete cases
incompleteRowCount <- sum(!complete.cases(data))
```

The total number of rows with NAs is: `r incompleteRowCount`  

The imputing strategy consists in replacing the NAs with the mean of the number
of steps taken in the same interval on all days.

```{r}
# Define a function for imputing a missing value with the mean of the
# values for the corresponding interval 
impute <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
# Impute the missing data
newData <- data %>%
  ddply(~ interval, transform, steps = impute(steps)) %>%
# Make sure the data is still ordered correctly
  arrange(date, interval)
# Calculate the sum of steps per day
stepsPerDay <- newData %>%
  group_by(date) %>%
  summarize(steps = sum(steps, na.rm = TRUE))
# Calculate the summary data (mean, median)
summaryAfterImpute = stepsPerDay %>%
  summarise(mean = round(mean(steps, na.rm = TRUE), 2),
            median = round(median(steps, na.rm = TRUE), 2))
# Plot the histogram
qplot(x = steps, data = stepsPerDay, geom = 'histogram')
```

The new mean of the total number of steps taken per day is: 
`r format(summaryAfterImpute$mean)`, compared to the previous 
`r format(summaryBeforeImpute$mean)`, a difference of 
`r format(abs(summaryAfterImpute$mean - summaryBeforeImpute$mean))`.  
The median of the total number of steps taken per day is: 
`r format(summaryAfterImpute$median)`, compared to the previous 
`r format(summaryBeforeImpute$median)`, a difference of 
`r format(abs(summaryAfterImpute$median - summaryBeforeImpute$median))`.  
As a result of imputing the missing data, the new mean and median are now equal.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Add a factor variable to differentiate between weekdays and weekends
newData <- newData %>%
  mutate(daytype = factor(!(wday(date) %in% 2:6),
                          labels = c('weekday', 'weekend')))
# Plot the time series
qplot(x = interval, y = steps, data = newData, geom = 'line', stat = 'summary',
      fun.y = 'mean') +
  facet_wrap(~ daytype, ncol = 1) +
  scale_x_datetime(breaks = '4 hours', labels = date_format('%H:%M'))
```
